Visual narratives of the COVID-19 pandemic

### 2022-10-27 JDSSV_manuscript_submitted.Rnw
final copyediting

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
   \And }
\Plainauthor{}

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
}

@@

Over the past two years, several waves of COVID-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience. According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and COVID-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. %So there are several goals to be achieved by visualising COVID-19 data.
These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.

@@

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.

@@

At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. 
These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}. 
Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
In other words, while our **technology** for recording, visually summarizing, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when **designing** charts which communicate data accurately and effectively.

@@

\subsection{Data visualisation Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualisation is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualisation (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

@@

\subsection{Scope}
Our focus in this paper is to highlight, analyze, and discuss a selection of charts and graphics which caught our attention during the initial, real-time experience of the pandemic. As we primarily consume media from western Europe and the United States, our assessment is primarily limited to news outlets catering to these audiences; a retrospective global analysis of COVID graphics would certainly be interesting, but is beyond the scope of this work. While the charts we have selected were of personal interest during the fear and uncertainty of the initial stages of the COVID pandemic, in this paper we assess these same charts retrospectively, from a world where COVID is an acknowledged fact of life. While it still has significant impacts on our day-to-day lives, the existence of treatments, vaccines, and other preventative measures, as well as the human tendency to adapt to changing circumstances mean that the novel coronavirus which emerged in 2019 no longer dominates our consciousness in the same way. Because of this perspective difference, we have enough distance from the initial cataclysm to be able to assess the emotional impact of the graphics at the time and to assess their utility retrospectively.

@@

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths during this initial period were commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. A central element of the narrative during March 2020 was the ubiquity of the disease and the accompanying global impact.

@@

Proportional symbol maps, see \Cref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight \citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as colour). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19 \citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
We will examine these topics more closely in \Cref{sec:rankings}.

@@

Throughout the pandemic, both in choropleth maps and time series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed \citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

@@

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. 
The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely. 
Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts \citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change \citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

@@

The most obvious time series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{ft-waves}\\
\includegraphics[width=\linewidth]{ft-waves-linear}
\caption{COVID Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom). Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
\label{fig:ft-waves}
\end{figure}

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube \citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names \citep{barryRemembering1000002020}. While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualisations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.

@@

If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates, however, we can gain some clarity as to their use in this particular context. We are abysmal at forecasting exponential growth, vastly under-estimating future growth by using linear or quadratic approximations \citep{wagenaarExtrapolationExponentialTime1978a,lawrenceExploringJudgementalForecasting1992,timmersInverseStatisticsMisperception1977a}. If the goal of a chart of COVID case counts is to allow individuals to forecast the trend and make decisions accordingly, then using a log scale (with all of its pitfalls in understanding) may be the best option, as it at least replaces the need to forecast along an exponential curve with the need to forecast along a straight line. However, it is not clear that individuals can transfer that prediction back to a linear scale. If the goal of presenting a chart of case counts is to tell the story of what has happened (rather than supporting future decision-making), then it is undoubtedly better to present the data on a more familiar linear scale that requires less cognitive load on the part of the viewer.

@@

\section{Summary and Discussion} \label{sec:summary}
Throughout the pandemic (thus far), visual representations of data have been an integral part of scientific communication. While not always optimal, these graphics have attempted to provide meaningful context, to encourage individual and collective action, and to help individuals grapple with the scale of the pandemic in cases, deaths, vaccinations, and interventions. As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic. This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.

The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes. Thus, analyzing the rich collection of visualisations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation. As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.

@@

The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualisation in the modern era. Even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge. It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.

@@

First ideas have been presented at DSSV 2020. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.

>>>>

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Susan VanderPlas\\
\affiliation{Statistics Department\\ University of Nebraska Lincoln\\ United States}
   \And Adalbert F.X. Wilhelm \\ \affiliation{School for Business\\ Social and Decision Sciences\\ Jacobs University Bremen}}
\Plainauthor{Susan VanderPlas, Adalbert F.X. Wilhelm}

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{Adalbert F.X. Wilhelm\\
School for Business, Social and Decision Sciences\\
Jacobs University Bremen\\
Campus Ring 1\\
28759 Bremen\\
Germany\\
e-mail:a.wilhelm@jacobs-university.de
}

@@

Over the past two years, several waves of COVID-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience. According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and COVID-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. %So there are several goals to be achieved by visualising COVID-19 data.
These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation, the general principles remain the same; ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.

@@

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles. Hence data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and has already become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.

@@

At the same time, we still see a lot of defective graphics disseminated and shared; some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. 
These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}. 
The use of numbers, their graphic representation and the systematic creation of charts and tables from data in the mass media have been accompanied from the beginning by efforts to improve the communication of quantitative and statistical information. And while efforts continue to be made in this area, historical documents describe many of the same problems we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
In other words, while our \textit{technology} for recording, visually summarising, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when \textit{designing} charts which communicate data accurately and effectively.

@@

\subsection{Data visualisation Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualisation is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: Which variables are shown, What limits are imposed, What visual forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualisation (computer, physical newspaper, tablet, cell phone) -- all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organisations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyse in this paper did not simply pop into being -- they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

@@

\subsection{Scope}
Our focus in this paper is to highlight, analyse, and discuss a selection of charts and graphics which caught our attention during the initial, real-time experience of the pandemic. As we primarily consume media from western Europe and the United States, our assessment is primarily limited to news outlets catering to these audiences; a retrospective global analysis of COVID graphics would certainly be interesting, but is beyond the scope of this work. While the charts we have selected were of personal interest during the fear and uncertainty of the initial stages of the COVID pandemic, in this paper we assess these same charts retrospectively, from a world where COVID is an acknowledged fact of life. While it still has significant impacts on our day-to-day lives, the existence of treatments, vaccines, and other preventative measures, as well as the human tendency to adapt to changing circumstances mean that the novel coronavirus which emerged in 2019 no longer dominates our consciousness in the same way. Because of this perspective difference, we have enough distance from the initial cataclysm to be able to assess the emotional impact of the graphics at the time and to assess their utility retrospectively.

@@

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths during this initial period were commonly visualised by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. A central element of the narrative during March 2020 was the ubiquity of the disease and the accompanying global impact.

@@

Proportional symbol maps, see \Cref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight \citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualising epidemic disease spread.
In addition, the visual impression of any map-based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as colour). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19 \citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organisations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram, and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe; none of the above visualisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: How are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
We will examine these topics more closely in \Cref{sec:rankings}.

@@

Throughout the pandemic, both in choropleth maps and time series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic, this situation was both extremely common and had real consequences; people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed \citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

@@

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex; over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. 
The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely. 
Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts \citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change \citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalisations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

@@

The most obvious time series plot is also the most common; plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalisations (and sometimes vaccinations), to provide a more comprehensive view of the situation. This is particularly useful as there has been some decoupling of case numbers and hospitalisations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{ft-waves}\\
\includegraphics[width=\linewidth]{ft-waves-linear}
\caption{COVID Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom). Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalisations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
\label{fig:ft-waves}
\end{figure}

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube \citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanise the names \citep{barryRemembering1000002020}. While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualisations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.

@@

If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates,  we can gain some clarity as to their use in this particular context. We are abysmal at forecasting exponential growth, vastly under-estimating future growth by using linear or quadratic approximations \citep{wagenaarExtrapolationExponentialTime1978a,lawrenceExploringJudgementalForecasting1992,timmersInverseStatisticsMisperception1977a}. If the goal of a chart of COVID case counts is to allow individuals to forecast the trend and make decisions accordingly, then using a log scale (with all of its pitfalls in understanding) may be the best option, as it at least replaces the need to forecast along an exponential curve with the need to forecast along a straight line. However, it is not clear that individuals can transfer that prediction back to a linear scale. If the goal of presenting a chart of case counts is to tell the story of what has happened (rather than supporting future decision-making), then it is undoubtedly better to present the data on a more familiar linear scale that requires less cognitive load on the part of the viewer.

@@

\section{Summary and Discussion} \label{sec:summary}
Throughout the pandemic (thus far), visual representations of data have been an integral part of scientific communication. While not always optimal, these graphics have attempted to provide meaningful context, to encourage individual and collective action, and to help individuals grapple with the scale of the pandemic in cases, deaths, vaccinations, and interventions. As in any developing situation, choices made at some points in the pandemic did not always persist -- many news outlets refined their charts and approach to the design of graphics over the course of the pandemic. This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.

The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes. Thus, analysing the rich collection of visualisations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation. As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good -- to educate and inform the public.

@@

The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualisation in the modern era. Even with numerous guidelines and standards for graphics, it is not straightforward to communicate complex data; creating charts which fully represent the intended message or messages remains a challenge. It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.

@@

First ideas have been presented at DSSV 2021. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
